Simultaneous Settler Projects and U.S. Racial State Formation

### 2024-06-11 ms_backend.Rmd
revision ff

---
title: "Settler colonial projects explain variation in US state formation in the early 20th c."
author: "Frank Edwards and Theresa Rocha Beardall"
date: "last updated `r Sys.Date()`"
output: bookdown::pdf_document2
---
```{r setup, include = F, message = F, cache = F}
knitr::opts_chunk$set(message = F, echo = F, warning = F)
library(tidyverse)
library(lme4)
library(maps)
library(usmap)
library(grid)
library(gridExtra)
library(brms)
library(tidycensus)
library(tigris)
library(units)
library(sf)
library(tidybayes)
library(rstan)
library(knitr)
library(trackdown)
trackdown::trackdown_auth_configure(path = "client_secret_243891325977-kgvbep6m6eb8t2feb8rvu9fneo8d6msa.apps.googleusercontent.com.json")
# trackdown::update_file(file = "ms.Rmd",
#                        hide_code = T,
#                        rich_text = T)
trackdown::download_file(file = "ms.Rmd")

options(tigris_use_cache = T)
options(scipen = 999)
options(dplyr.print_max = 30)
map<-map_data("state")
theme_set(theme_minimal())
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
# set hook for thousands comma
knit_hooks$set(inline = function(x) {
prettyNum(round(x,2), big.mark=",")
})
```

@@

```{r run-models, cache = T, include = F, message = F}
source("models.R")
```

# Introduction

Settler colonialism deeply and unevenly impacts the development of social policy systems. Settler colonial regimes engage in diverse forms of dispossession, extraction, exploitation, and social reproduction. We show that spatial and temporal variation in the salience of settler projects is closely related to the kinds of infrastructure that settler colonial states develop for social control and social reproduction. We show that the formation of the social state in the early 20th century in the US varied dramatically across places as a function of geographic variation in the salience of a set of interests of the settler colonial project: land dispossession, labor exploitation, and reproduction of the settler polity.

We examine three deeply intertwined, but spatially and temporally variable, contexts for local policymaking in the Progressive Era (1900 - 1930): the Jim Crow agricultural economy; mass urbanization and industrialization; and the rapid distribution of Native lands to white settlers. Each of these contexts presented different sets of problems to local and national settler policy elites.

Under sharecropping regimes in the US South, white settler elites sought to create and reproduce brutal and paternalistic regimes of informal and formal social regulation that kept Black workers and their families dependent on and subordinate to white interests. Industrial and social white elites in the North sought to discipline large numbers of European immigrants toward nascent ideals of a white American settler polity oriented toward a narrow gendered division of labor and subordination to the interests of industrial capital. In 1862, the federal government enacted the Morill, Homestead, and Pacific Railroad Acts, each of which provided mechanisms for the rapid distribution of dispossessed Native lands to settlers. White settlers and settler corporate interests flooded recently dispossessed Native lands with the promise of free or cheap title, seeking to rapidly develop the settler nation in recently admitted states.

We suggest that each of these contexts yielded different trajectories for the formation of US state infrastructure. In the analysis that follows, we leverage data from a range of archival sources to map contexts for early 20th century settler colonization across the US, then build statistical models to evaluate whether these contexts were related to the kinds of capacities that local governments developed for contemporaneous social control and social reproduction.

The foundations upon which states built out their infrastructure to control, educate, and care for their populations were closely related to regional variation in the political economy of settler colonialism across regions in the US. States developed more or less robust institutions in various domains to address the problems of governance posed by the local projects of settler colonization that both elites and ordinary settlers sought to advance.

We focus our attention on three interconnected settler colonial structures and processes that had large impacts on social life in the early 20th century US: 1) the active dispossession and erasure of Native nations and the development of settler enterprises and communities; 2) the exploitation of Black labor for agricultural production and the fortification of an apartheid social order; and 3) the coercive and paternalistic assimilation of recent immigrants from Eastern and Southern Europe into an urban industrial settler colonial polity. We argue that these deeply related, but geographically uneven, processes were related to the development of state capacities to resolve emerging problems of domestic governance in a period of rapid social transformation. In more general terms, we argue that the local salience and configuration of key social structures, like settler colonization, are closely related to local trajectories for state formation and institution building.

Consistent with the historiography on the Progressive Era (GORDON, FOX, OTHERS), we find clear relationships between place and policy infrastructure. State formation trajectories varied dramatically across the US, and this variation is closely related to the local salience of distinct racialized settler colonial projects. Our work bridges these sometimes disparate historical accounts by showing that the local interests of settler elites were closely tied to the kinds of social and racial projects that the early carceral and welfare states engaged in.

# Three settler colonial projects in the early 20th c. US

The geographic distribution of the local implementation of the three focal settler projects we focus on is crudely illustrated in Figure \@ref(fig:three-maps). The federal government was most heavily involved in actively redistributing Native lands to settlers in states in the Midwest, Great Plains, Mountain West, and on the West Coast. While Native land dispossession was still deeply salient in Eastern states, questions of settler land ownership and the establishment of coherent settler polities had been resolved (albeit incompletely) prior to the late 19th and early 20th centuries. In panel 1, we show the number of acres that had been dispossessed from Native nations and distributed to settlers and settler institutions through two key laws that heavily impacted patterns of colonial settlement in the late 19th and early 20th centuries, the Morill and Homestead acts of 1862.

```{r three-maps, fig.cap = "Three focal contexts for settler colonial governance, 1910"}
### ADD DISPOSESSION
p1 <- dist %>% 
  select(STUSPS, year,
         morill_acres_cum, homestead_acres_cum,
         total_acres_2021) %>% 
  filter(year == 1910) %>% 
  mutate(dist = morill_acres_cum + homestead_acres_cum) %>% 
  left_join(st) %>% 
  ggplot(aes(fill = dist / 1e6,
             geometry = geometry)) + 
  geom_sf() + 
  theme_void() + 
  scale_fill_viridis_c() + 
  theme(legend.position = "bottom") + 
  labs(fill = "Acres\n(millions)",
       subtitle = "1. Lands patented through\nMorill and Homestead acts")
p2 <- st %>% 
  left_join(pop %>% 
              filter(year == 1910)) %>% 
  ggplot(aes(fill = blk / tot_pop)) + 
  geom_sf() + 
  theme_void() + 
  scale_fill_viridis_c() + 
  theme(legend.position = "bottom") + 
  labs(fill = "Proportion",
       subtitle = "2. Black population")
p3 <- st %>% 
  left_join(pop %>% 
              filter(year == 1910)) %>% 
  ggplot(aes(fill = urban_immig_pop / tot_pop)) + 
  geom_sf() + 
  theme_void() + 
  scale_fill_viridis_c() + 
  theme(legend.position = "bottom") + 
  labs(fill = "Proportion",
       subtitle = "3. Urban immigrant population")
grid.arrange(p1, p2, p3, nrow = 1)
```

The Jim Crow system of apartheid was developed and became deeply entrenched across the states in the US South that had built elaborate settler plantation economies through the exploitation and terrorization of formerly enslaved Black people. As of 1910, the first wave of the Great Migration of African Americans out of the US South was underway, but, as shown in Panel 2 of Figure \@ref(fig:three-maps), the Black population in the US was still heavily concentrated in the US Southeast.

And the Industrial Revolution, coupled with a wave of crises across Europe, created both strong push and pull forces for large populations of immigrants to travel to North America. These immigrants settled across the industrial cities of the Northeast, Midwest, and West coasts. Many of them also joined the waves of white settlers pursuing the opportunity to claim recently dispossessed Native lands in the Central and Western portions of the US. As shown in Panel 3 of Figure \@ref(fig:three-maps), the populations of many states in the Northeast, Midwest, and West Coast were becoming increasingly urban and foreign born.

## Building the settler nation-state

KEY CHARARCTERISTICS: RECENT OR ONGOING DISPOSESSION / REMOVAL, RECENTLY ADMITTED STATES, RELATIVELY SPARSE WHITE POPS. ELIMINATION / REPLACEMENT AS KEY DYNAMIC.

In places where land speculation and homesteading fueled rapid white settlement, governments and churches hired large numbers of teachers for the intertwined goals of building a white settler polity on the 'frontier' and to continue the coerced assimilation of Indigenous children through a notorious and brutal system of boarding schools.

```{r cessions-map, fig.cap = "Land cessions from Native nations claimed by US government 1795 - 1893"}
st<-states(cb = T) %>%
  st_transform(8528) %>% 
  filter(!STUSPS%in%
           c("HI", "AK", 
             "PR", "GU", "AS","VI",
             "MP"))
st_acres<-data.frame(state = st$STUSPS,
                     total_acres_2021 = as.numeric(set_units(st_area(st), "acre")))
### royce map data
## available here: https://data-usfs.hub.arcgis.com/datasets/usfs::tribal-lands-ceded-to-the-united-states-feature-layer/about
## and archived here: https://data.nativeland.info/dataset/indian-land-cessions-in-the-united-states-1784-1894/resource/e2b874d5-3c39-4287-bf89-d4ef14b87c83
cessions<-read_sf("./data/us_forest_cessions_royce/S_USA.TRIBALCEDEDLANDS.shp")
cessions_dat<-read_csv("./data/us_forest_cessions_royce/royce.csv") %>% 
  mutate(time = as_date(TribalCededLandsTableCessDate1),
         year = year(time)) %>% 
  rename(CESSNUM = TribalCededLandsNewCESSNUM) %>% 
  select(CESSNUM, year)
cessions<-cessions %>% 
  left_join(cessions_dat)
cessions<-cessions %>% 
  st_transform(8528)
cessions_diff<-st_difference(st, st_union(cessions))
cessions_diff<-cessions_diff %>% 
  mutate(year = NA)
### layer state boundaries
cessions<-st_intersection(cessions, st)
### add in settler claimed territories as of 1796
cessions2 <- cessions %>%  
  bind_rows(cessions_diff %>%  
              mutate(year = NA))
ggplot(cessions2,
       aes(fill = year)) + 
  geom_sf() + 
  theme_void() + 
  scale_fill_viridis_c() + 
  labs(fill = "Year")

Figures \@ref(fig:cessions-map) and \@ref(fig:cessions-ts) show the temporal and spatial process of the dispossession of lands from Native Nations by the US Government between 1795 and 1893 (DATA FROM USFS VIA Lib Of Cong). These maps show a clear difference in the timing of land cessions as a function of place. By 1795, the US Government claimed nearly all lands East of the Appalachian Mountains, including large parts of contemporary Kentucky and Tennessee. Land dispossession throughout the South and the Great Lakes regions accelerated quickly during the early 19th century. Throughout the middle of the 19th century, there were large-scale efforts by the federal government and vigilante settlers to dispossess tribal lands in California, the Midwest, Great Plains, and Mountain West. By 1895, nearly all lands in the contemporary lower 48 US states had been claimed by the federal government as US territory.

```{r cessions-ts, fig.cap = "Cumulative disposession of Native lands in the 19th c. by US region and state as proportion of contemporary state area. Each line represents one state."}
# p1<-ggplot(disp,
#        aes(x = year,
#            group = STUSPS)) + 
#   geom_line(aes(y = cumulative_ceded / 1e6)) + 
#   labs(x = "Year",
#        y = "Acres (millions)") + 
#   lims(y = c(0, 150)) + 
#   facet_wrap(region~division) + 
#   labs(subtitle = "1. Total area")
p2<-ggplot(disp %>% 
             filter(year > 1795),
       aes(x = year,
           group = STUSPS)) + 
  geom_line(aes(y = prop_ceded)) + 
  labs(x = "Year",
       y = "Proportion of state area") + 
  facet_wrap(region~division) 
p2
```

The Federal government passed a series of laws in 1862 designed to rapidly redistribute dispossessed Native lands to white settlers and to large corporate and state institutions. The Homestead, Morill, and Pacific Railway Acts of 1862 provided legal frameworks through which expropriated Native lands could be rapidly redistributed to white settlers, large corporations, and Universities. Figure \@ref(fig:hs-morill-map) shows the total number of acres within a state that had been granted to settlers and settler institutions by 1930. More than 10 million acres of Native lands had been redistributed to US land grant Universities at little or no cost for their use or liquidation by 1930 under the Morill act. Under the Homestead Act, more than 255 million acres of Native land had been transferred to settlers by 1930. States in the Midwest, Great Plains, Mountain West, and West Coast saw the highest volumes of land redistribution under these laws.

```{r hs-morill-map, fig.cap = "Lands patented under two legislative mechanisms for the redistribution of disposessed Native lands to settlers, cumulative total 1862 - 1930"}
pdat<-disp %>% 
  select(STUSPS, total_acres_2021) %>% 
  distinct() %>% 
  left_join(dist %>% 
              filter(year == 1930) %>% 
              select(STUSPS, morill_acres_cum, homestead_acres_cum)) %>% 
  replace_na(list(morill_acres_cum = 0, homestead_acres_cum = 0)) %>% 
  left_join(st)
p1<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = morill_acres_cum / 1e6)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Acres (millions)",
       subtitle = "1. Morill Act") + 
  theme(legend.position = "bottom")

p2<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = homestead_acres_cum / 1e6)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Acres (millions)",
       subtitle = "2. Homestead Act") + 
  theme(legend.position = "bottom")
grid.arrange(p1, p2, ncol = 2)
```

## Jim Crow

KEY CHARACTERISTICS: ANTI-BLACK SOCIAL, LEGAL, ECONOMIC REGIME. WHITE SETTLER HOMOGENEITY

By 1877 efforts to build effective racial democracy through post-Civil War Reconstruction were abandoned by the federal government. By the turn of the century, white elites had legislated and entrenched a regime of brutal racial exploitation and apartheid across the US South (DU BOIS, FONER, FRANKLIN, CURTIN).

DEFINE JIM CROW REGIME. APARTHEID, FORMAL / INFORMAL TERROR REGIME, BLACK LABOR EXPLOITATION, PRIMARILY AG through SHARECROPPING.

READ APD CITATIONS FROM ANN REVIEWS ON JIM CROW, APD, RACISM. DOES APD SUPPORT THE WEAK STATE THESIS?

In places where tenant farming by Black families was common, there was relatively little investment in state bureaucracies. Social control and reproduction instead relied on a coordinated formal and informal system of racial terrorism and apartheid that developed alongside a relatively weak set of state social policy institutions (QUADAGNO, KATZNELSON).

Figure \@ref(fig:jim-crow-maps) shows two key demographic features of this period. In panel 1, we illustrate the entrenchment and ubiquity of the sharecropping system of racially exploitative agricultural production. In panel 2, we show that, unlike much of the rest of the United States at the time, the white population of the US South was composed overwhelmingly of white settlers who had been born in the United States.

```{r jim-crow-maps, fig.cap = "Black tenant farming and immigrant status of white settlers, 1910"}
pdat<-st %>% 
  left_join(pop) %>% 
  filter(year == 1910)
p1<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = blk_sharecrop / 1e5)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Population\n(hundred thousands)",
       subtitle = "1. Black population living on tenant farms") + 
  theme(legend.position = "bottom")
p2<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = white_immig / (white_immig + white_non_immig))) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Proportion",
        subtitle = "2. Proportion of white population foreign born") + 
  theme(legend.position = "bottom")
grid.arrange(p1, p2, nrow = 1)
```

## Regulating and incorporating urban white immigrants

KEY CHARACTERISTICS: URBANIZATION, INDUSTRIALIZATION, MORAL PANICS OVER IMMIGRANT CULTURAL DIFFERENCE. (TWO ROUTES, REACTIONARY DISCIPLINE, LIBERAL WELFARE PATERNALISM, BOTH ACTIVE).

In the rapidly urbanizing portions of the Northeast, Midwest, and West Coast, we see heavy investment in what has become the ideal type of the Progressive social state: a robust and professionalizing police force and a broad paternalistic welfare state. These institutions were developed by elites, in part, to regulate urban disorder discursively linked to growing populations of European immigrants and Black migrants from the US South, and with rapidly assimilating these groups into the norms and ideologies of industrial capitalism and the settler nation.

The industrialization of the US Northeast and Midwest during the late 19th and early 20th centuries spurred two related demographic processes: rapid urbanization and mass European immigration. Figure \@ref(fig:immig-maps) displays the distribution of both the urban population and the white immigrant population in the US in 1910. States like New York, Massachusetts, Pennsylvania, Ohio, and Illinois saw dramatic growth in their populations, with much of that growth composed of European immigrants and concentrated in cities like Chicago, Boston, Cleveland, Philadelphia, and New York.

```{r immig-maps, fig.cap = "White immigrant population and urban population, 1910"}
pdat<-st %>% 
  left_join(pop) %>% 
  filter(year == 1910)
p1<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = white_immig / 1e6)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Population\n(millions)",
       subtitle = "1. White, foreign-born") + 
  theme(legend.position = "bottom")
p2<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = urban_pop / 1e6)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Population\n(millions)",
        subtitle = "2. Urban") + 
  theme(legend.position = "bottom")
grid.arrange(p1, p2, nrow = 1)
```

Progressive reformers in the Northeast and Midwest were preoccupied with solving a battery of social problems associated with urban life, industrial capitalism, and their perceptions of the culture and behavior of Southern and Eastern European immigrants.

HIGH LEVEL SUMMARY OF 1st WAVE WHITE FEMINISM AND SOCIAL WORK VIA GORDON, HARTMAN, OTHERS.

The emerging fields of sociology, social work, and criminology also sought to apply scientific principles to the organization and practice of policing and corrections. HIGH LEVEL SUMMARY OF PROG ERA POLICING, JUV COURT, PRISON REFORM. LINK TO GOTTSCHALK ON CARCERAL STATE EMERGENCE FROM THESE MOVEMENTS.

# Patterns in the development of social policy infrastructure 1900 - 1930

We define social state capacity as the resources available to governments to direct and intervene in processes of social control and social reproduction (see Cingolani 2013 for a detailed review of theory and measurement). Capacity is a function of both physical infrastructure and human resources available to state actors. We focus on a broad set of social policy systems including education, social welfare, and criminal justice.

We rely on three sets of measures associated with the capacity of states to accomplish social policy goals: 1) employment rates in key street-level social policy positions as a measure of the relative size of bureaucracies and magnitude of human effort available to policy actors, 2) populations housed in institutional group quarters as a measure of the scope of particular kinds of physical infrastructure used by the welfare and carceral systems during the Progressive Era, and 3) counts of operating Indian boarding schools as indicators of physical infrastructure directed at coerced assimilation and erasure of Indigenous peoples. Taken together, these three sets of measures provide a diverse and broad set of insights into spending on the basic infrastructure of social governance during the early 20th century.

## Employment in education, policing, and social welfare: 1900 - 1930

There was dramatic growth in public sector employment in general (CLEAR FROM FED GOVT EMPLOYMENT, OTHER CITATIONS), and in roles targeted at nation building, social control, and social welfare in particular during the Progressive Era. *These dramatic increases in employment rates were reflected in the increasing complexity and reach of a wide range of public institutions described by a number of historians (GENERAL LIT ON GOVT COMPLEXITY AND ARTICULATION 1900 - 1930).*

In Figure \@ref(fig:emp-ts), we display employment rates for the lower 48 US States between 1900 and 1930 in panel 1, and the proportional change in employment rates by category over time in panel 2. Employment of teachers increased from a rate of 5.5 per 1,000 population in 1900 to a rate of 8.6 per 1,000 in 1930, a 56 percent increase. Employment of police doubled during this period, increasing from a rate of about 0.7 per 1,000 in 1900 to a rate of about 1.4 per 1,000 in 1930. Outside of a few cities in the northeast, social workers were almost never recorded in the Census in 1900 (only 361 recorded nationally in 1900). However, by 1930, the Census recorded about 17,000 social workers nationally, an increase of about 30 fold in terms of employment rates per capita. CHECK THESE NUMBERS

```{r emp-ts, fig.cap = "Employment in three core social policy professions over time, US rates per 1,000 population, 1900 - 1930"}
emp_pdat<-pop %>% 
  group_by(year) %>% 
  summarize(pop = sum(tot_pop),
            Teachers = sum(teachers),
            Police = sum(police),
            social_workers = sum(socwork)) %>% 
  pivot_longer(cols = Teachers:social_workers) %>% 
  mutate(name = case_when(
    name == "Teachers"~ "Teachers",
    name == "Police" ~ "Police",
    name == "social_workers" ~ "Social Workers")) %>% 
  mutate(rate = value / pop * 1e3)
### convert to proportional change
emp_pdat00<-emp_pdat %>% 
  filter(year == 1900) %>% 
  rename(rate00 = rate) %>% 
  select(-year, -value, -pop)
emp_pdat_prop <- emp_pdat %>% 
  left_join(emp_pdat00) %>% 
  mutate(prop_change = rate / rate00) 
p1<-emp_pdat_prop  %>% 
  ggplot(aes(x = year, y = rate, color = name)) + 
  geom_point() + 
  geom_line() + 
  labs(x = "Year", y = "Rate per 1,000",
       color = "",
       subtitle = "1. Employment rate") + 
  theme(legend.position = "bottom")
p2<-emp_pdat_prop  %>% 
  ggplot(aes(x = year, y = prop_change, color = name)) + 
  geom_point() + 
  geom_line() + 
  labs(x = "Year", y = "Proportional change (log scale)",
       color = "",
       subtitle = "2. Change relative to 1900") + 
  scale_y_log10() + 
  theme(legend.position = "bottom")

grid.arrange(p1, p2, nrow = 1)
```

However, this expansion of the reach of state and state-aligned street-level bureaucrats into social life did not occur uniformly across the country. Figure \@ref(fig:emp-maps) displays employment rates per 1,000 population in 1930 for teachers, police, and social workers for each of the lower 48 US States. Employment rates for teachers were lowest-in-the nation in the US South. Employment rates for teachers were highest in a cluster of Great Plains states, like the Dakotas, Wyoming, Nebraska, Iowa, and Kansas. Employment rates for police were at their peak in 1930 in Northeastern, Midwestern, and West Coast states like New York, New Jersey, Massachusetts, Illinois, and California. Police employment rates were relatively lower in places like Arkansas, Mississippi, and North Carolina. The distribution of social workers in 1930 shares some features with the distribution of police: states in the Northeast, Midwest, and West coast like New York, Ohio, Massachusetts, California, and Minnesota had high rates of social worker employment, while states in the Great Plains and South, like Mississippi and South Dakota recorded very few social workers in the 1930 Census.

```{r emp-maps, fig.cap = "Employment in three core social policy professions across US states in 1930, rates per 1,000 population"}
pdat<-st %>% 
  left_join(pop) %>% 
  filter(year == 1930)
p1<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = teachers / tot_pop * 1e3)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Rate",
       subtitle = "1. Teachers") + 
  theme(legend.position = "bottom")
p2<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = police / tot_pop * 1e3)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Rate",
        subtitle = "2. Police") + 
  theme(legend.position = "bottom")
p3<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = socwork / tot_pop * 1e3)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Rate",
        subtitle = "3. Social Workers") + 
  theme(legend.position = "bottom")
grid.arrange(p1, p2, p3, nrow = 1)

## Physical infrastructure: prisons, jails, asylums, orphanages and poorhouses

The capacity of state welfare and carceral systems is also a function of their ability to engage in therapeutic, correctional, or paternalistic confinement of socially problematic populations. In Figure \@ref(fig:gq-map) we show the distribution of the population housed in institutional group quarters in 1930 across US states. Using the 1930 Census, we display the population housed in prisons in jails in panel 1, the population housed in mental institutions in panel 2, and the population housed in institutions for the poor and/or disabled in figure 3. Note that data quality issues in the 1910 Census limit the possibilities for time series analysis.

Panel 3 displays the rate of confinement in institutions for the poor and disabled across US states in 1930. These institutions included poorhouses, settlement homes, orphan asylums, sanitariums, and similar institutions targeted at containing and/or supporting social groups understood to be 'dependent' on state or religious charity support. We find that rates of confinement (likely closely correlated with capacity) were lowest in states in the South, like Arkansas, Alabama, Mississippi and Florida. New England, the mid-Atlantic, and Great Lakes states had relatively high rates of confinement in these kinds of institutions. The East North Central Division, including states like Wisconsin, Illinois, and Ohio had the highest average rate of confinement, at 4.3 per 1,000 persons, while the West South Central division (Arkansas, Louisiana, Oklahoma, and Texas) had the lowest rate of confinement at 1.4 per 1,000 persons in 1930.

Panel 2 displays regional patterns in confinement in institutions for the mentally ill across places in 1930. States in the mid Atlantic, New England, and West Coast had the highest rates of confinement, at over 3 per 1,000 on average. In New Hampshire and Massachusetts, confinement rates in 1930  exceeded 4 per 1,000 population. States in the East South Central Division (Mississippi, Alabama, Tennessee, and Kentucky) had the lowest average levels of confinement, at 1.4 per 1,000 persons. Levels of confinement ranged between 2 and 3 on average across the rest of the country. Of note, we observe noinmates in mental institutitons in North Dakota in the 1930 census data.

Whereas states in the South tended to have very low levels of confinement in institutions for the poor and disabled and relatively low levels of confinement in mental institutions, they had levels of confinement in correctional institutions that were near national averages. This suggests that Southern policy makers and philanthropic allies systematically invested fewer resources in physical infrastructure targeted at containing or supporting the poor, young, mentally ill, and disabled than did other states, while maintaining and developing prisons and jails at rates similar to the rest of the nation.

```{r gq-map, fig.cap = "Population confined in prisons, mental institutions, and institutions for the poor and disabled, 1930"}
pdat<-st %>% 
  left_join(pop) %>% 
  left_join(bs) %>% 
  filter(year == 1930) %>% 
  mutate(inst_correctional_pc = inst_correctional / tot_pop * 1e3,
         inst_mental_pc = inst_mental / tot_pop * 1e3,
         inst_poor_disabled_pc = inst_poor_disabled / tot_pop * 1e3)

p1<-ggplot(pdat) +
  geom_sf(aes(geometry = geometry,
              fill = inst_correctional_pc)) +
  scale_fill_viridis_c() +
  theme_void() +
  labs(fill = "Rate",
       subtitle = "1. Correctional institutions") +
  theme(legend.position = "bottom")
p2<-ggplot(pdat) +
  geom_sf(aes(geometry = geometry,
              fill = inst_mental_pc)) +
  scale_fill_viridis_c() +
  theme_void() +
  labs(fill = "Rate",
        subtitle = "2. Mental institutions") +
  theme(legend.position = "bottom")
p3<-ggplot(pdat) +
  geom_sf(aes(geometry = geometry,
              fill = inst_poor_disabled_pc)) +
  scale_fill_viridis_c() +
  theme_void() +
  labs(fill = "Rate",
        subtitle = "3. Institutions for the\npoor and disabled") +
  theme(legend.position = "bottom")
grid.arrange(p1, p2, p3, nrow = 1)

# for in-text callouts
pdat<-st %>% 
  left_join(pop) %>% 
  left_join(bs) %>% 
  filter(year == 1930)%>% 
  mutate(inst_correctional_pc = inst_correctional/tot_pop * 1e3,
         inst_mental_pc = inst_mental / tot_pop * 1e3,
         inst_poor_disabled_pc = inst_poor_disabled / tot_pop * 1e3)

pdat_callout<-pdat %>% 
  pivot_longer(cols = inst_correctional_pc:inst_poor_disabled_pc,
               names_to = "inst_type",
               values_to = "rate") %>% 
  group_by(inst_type) %>% 
  summarize(mean = mean(rate),
            sd = sd(rate)) %>% 
  as_tibble() %>% 
  select(-geometry)
```

## Physical infrastructure: Indian boarding schools

The Indian boarding school epitomizes the sustained effort by the US government and its allies in settler religious and philanthropic organizations to eliminate Native nations through the coercive assimilation of Indigenous children. Investment in Indian boarding schools peaked during the Assimilation and Allotment era of Federal Indian Law, roughly spanning 1887 - 1934. Notably, these carceral and genocidal institutions represented an early, substantial, and sustained investment by the federal government in education infrastructure.

Below, we provide a descriptive overview of the geography and timing of boarding school operations. Following initiatives by Secretary of the Interior Deb Haaland and leaders from across Indian Country, the US Department of the Interior has initiated its first comprehensive review of the scope and impacts of the Indian boarding school program. We rely on data produced by the Department of the Interior (NEWLAND 2022) for historical counts of boarding schools in operation by state and year. We pair these data with information coded by our research team from archived annual reports produced by the Commissioner on Indian Affairs between 1890 and 1920 (CITATIONS).

```{r bsmap, fig.cap = "Boarding schools in operation in 1907 by state"}
bs_ts<-agents %>% 
  left_join(bs) %>% 
  left_join(st)

ggplot(bs_ts %>% 
         filter(year == 1907),
       aes(fill = boarding_schools,
           geometry = geometry)) + 
  geom_sf() + 
  scale_fill_viridis_c(trans = "sqrt") + 
  theme_void() + 
  labs(fill = "Facilities\noperating") + 
  theme(legend.position = "bottom")
```

@@

To explore the relationship between settler colonial political and economic contexts and state formation, we evaluate whether a set of measures of policy infrastructure are related to a set of measures that are related to locally-specific settler colonial policy contexts at the state and year-levels. We conduct exploratory and inferential analysis using data on US states between 1900 and 1930.

@@

\[E[\textrm{infrastructure}_{st}] = f(\textrm{settler interests}_{st})\]

We proceed by producing and evaluating descriptive data visuals for our operational measures of settler colonial policy contexts and comparing distributions of these measures across states and over time. We then produce and evaluate a similar set of visuals to evaluate the distribution of our social policy infrastructure measures. Finally, we estimate regression models that evaluate the relationships between our measures of infrastructure and settler interests, conditional on time-stable place variation, national temporal trends, and basic characteristics of the population in the state and year.

@@

*Equation 1: Model for count outcomes*

\[ y_{st} \sim \textrm{NegBin}(\lambda, \alpha)\]

\[ \lambda = \textrm{E}[y_{st}] \]

>>>>

---
title: "Simultaneous Settler Projects and U.S. Racial State Formation
METHODS AND FINDINGS"
author: "Frank Edwards and Theresa Rocha Beardall"
date: "last updated `r Sys.Date()`"
output: bookdown::pdf_document2
---
```{r setup, include = F, message = F, cache = F}
knitr::opts_chunk$set(message = F, echo = F, warning = F)
library(tidyverse)
library(lme4)
library(maps)
library(usmap)
library(grid)
library(gridExtra)
library(brms)
library(tidycensus)
library(tigris)
library(units)
library(sf)
library(tidybayes)
library(rstan)
library(knitr)
# library(trackdown)
# # trackdown::trackdown_auth_configure(path = "client_secret_243891325977-kgvbep6m6eb8t2feb8rvu9fneo8d6msa.apps.googleusercontent.com.json")
# # trackdown::update_file(file = "ms.Rmd", 
# #                        hide_code = T,
# #                        rich_text = T)
options(tigris_use_cache = T)
options(scipen = 999)
options(dplyr.print_max = 30)
map<-map_data("state")
theme_set(theme_minimal())
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
# set hook for thousands comma
knit_hooks$set(inline = function(x) {
prettyNum(round(x,2), big.mark=",")
})
```

@@

```{r run-models, cache = T, include = F, message = F}
# source("models.R")
mdat<-pop %>% 
  left_join(dist) %>% 
  left_join(bs) %>% 
  mutate(blk_sharecrop_pc = blk_sharecrop / tot_pop,
         urban_immig_pc = urban_immig_pop / tot_pop,
         homestead_plus_morill = homestead_acres_cum + morill_acres_cum)

#### set weakly informative priors for model parameters
### for m1
### try a growth model
mdat<-mdat %>% 
  mutate(year_c = (year - 1900)/10)

teachers_m1<-readRDS("./models/teachers_m1.rds")
cops_m1<-readRDS("./models/cops_m1.rds")
socwork_m1<-readRDS("./models/socwork_m1.rds")
prison_m1<-readRDS("./models/prison_m1.rds")
inst_m1<-readRDS("./models/inst_m1.rds")
bs_m1_z<-readRDS("./models/bs_m1_z.rds")
bs_m1<-readRDS("./models/bs_m1.rds")
```

@@

We describe the relationship between settler colonial political and economic contexts and state formation by evaluating whether a set of measures of policy infrastructure are related to a set of measures that are related to locally-specific settler colonial policy contexts at the state and year-levels. We use data on US states between 1900 and 1930 derived from US Census and archival sources.

Our approach is theory-driven, informed by 1) RACIAL STATE THEORY and 2) the HISTORIOGRAPHY OF EACH REGION IN THIS PERIOD. We suggest that RACIAL STATE THEORY provides a coherent framework for understanding heterogeneity in state formation during this period, by applying the simultaneous settler projects framework to these distinct contexts.

@@

$E[\textrm{infrastructure}_{st}] = f(\textrm{settler interests}_{st})$

We prioritize a descriptive approach that illustrates differences in macro-social contexts across places and their associations with state-formation outcomes and trajectories. We describe and evaluate the geographic and temporal co-occurrence of state formation outcomes and discrete settler projects.

*While there have been a number of compelling analyses of the long-run effects of variation in colonial systems on colonized societies (ACEMOGLU ETC) and of chattel slavery and racial terror on contemporary politics (CITATIONS) we refrain from engaging in formal causal inference in this analysis. Rich theoretical and historical work suggests that racial and settler colonial processes deeply structure policy outcomes through complex and long-running processes that become obscured and entangled with a battery of social processes. We suggest that these processes are visible at a relatively high-level in the historical data. However, settler colonialism is endogenous to the whole of US politics, economics, and demography.* TIGHTEN THIS

We proceed by producing and evaluating descriptive data visuals and maps of our operational measures of settler colonial policy contexts and comparing distributions of these measures across states and over time. We then produce and evaluate a similar set of visuals to evaluate the distribution of our social policy infrastructure measures. We pay particular attention to regional patterns in these distributions, and situate our interpretations in dialogue with prior historical and theoretical research.

Finally, we estimate regression models that evaluate the relationships between our measures of infrastructure and settler interests, conditional on time-stable place variation, national and regional temporal trends, and basic characteristics of the population in the state and year. These models formally evaluate associations between the geographies of settler colonial projects and a range of policy outcomes.

@@

**(1) Regression model for count outcomes**

$y \sim \textrm{NegBin}(\lambda, \alpha)$

$\lambda = \textrm{E}[y]$

$\log \left( \frac{\textrm{E}[y]}{\textrm{population}} \right) = \gamma_s + \beta \textrm{X}_{st}$

$x_{st} = t \times \left(\frac{\textrm{Black tenant farmers}_{st}}{\textrm{population}_{st}} +  \frac{\textrm{White urban immigrants}_{st}}{\textrm{population}_{st}} +\textrm{Morill and Homestead acres}_{st}\right)$

$\gamma \sim N(0, \sigma^2_\gamma)$
